---
title: "Choosing a transformation: Box-Cox in practice"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing a transformation: Box-Cox in practice}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4.2)
```

```{r setup, message = FALSE}
library(shewhartr)
library(ggplot2)
```

Box & Cox (1964) introduced a one-parameter family of power
transformations,

$$
y(\lambda) =
\begin{cases}
(x^\lambda - 1)/\lambda & \lambda \neq 0 \\
\log(x) & \lambda = 0,
\end{cases}
$$

and a procedure for choosing $\lambda$ by maximum likelihood. The
goal is to find a scale on which the residuals are approximately
normal and homoscedastic — the assumptions that classical
inferential tools, including Shewhart charts, presuppose.

`shewhart_box_cox()` returns the profile log-likelihood, the
maximiser $\hat \lambda$, and a 95% confidence interval based on the
chi-square approximation to twice the log-likelihood drop.

## A textbook example

```{r}
set.seed(2025)
y <- rlnorm(200, meanlog = 0, sdlog = 0.5)   # log-normal -> lambda = 0
bc <- shewhart_box_cox(y)
bc
```

The optimal lambda is near zero (log transformation), and the 95% CI
should cover zero. Let's plot the profile:

```{r, eval = FALSE}
autoplot(bc)
```

If the CI for $\lambda$ contains 1, no transformation is needed (the
data are approximately normal as is). If it contains 0, take logs.
If it contains 0.5, take square roots — and so on.

## Interaction with `shewhart_regression(model = "auto")`

The `"auto"` model in `shewhart_regression()` calls
`shewhart_box_cox()` internally on the response (with a +1 shift to
keep zeros valid) and selects among `linear`, `log`, `loglog`
according to the value of $\hat \lambda$:

* $|\hat \lambda - 1| \le 0.1$ → `linear`
* $|\hat \lambda - 0| \le 0.1$ → `log`
* $|\hat \lambda - 0.5| \le 0.1$ → `loglog`
* otherwise default to `linear` with a warning

This is a guidance step, not a guarantee. Always inspect the
residual diagnostics afterwards via `shewhart_diagnostics()`.

## When *not* to transform

If your data are counts, proportions, times-to-event, or other
quantities with a known parametric family, model that family
explicitly. Box was clear about this: if you can model the right
distribution, do so. Transforms exist for the case where the right
distribution isn't tractable and a normal approximation on a
suitably-chosen scale is the best available compromise.

The c, u, p, and np charts in this package implement that advice:
they support `limits = "poisson"` (or `"binomial"`) for exact
distribution-aware limits, instead of relying on a transformation
to coerce counts into approximate normality.

## References

- Box, G. E. P., & Cox, D. R. (1964). An Analysis of Transformations.
  *Journal of the Royal Statistical Society B*, 26(2), 211-252.
- Atkinson, A. C. (1985). *Plots, Transformations and Regression*.
  Oxford.
- Box, G. E. P., Hunter, W. G., & Hunter, J. S. (2005). *Statistics
  for Experimenters* (2nd ed.). Wiley.
- Sakia, R. M. (1992). The Box-Cox Transformation Technique: A Review.
  *The Statistician*, 41(2), 169-178.
