---
title: "Regression-based control charts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Regression-based control charts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4.2
)
```

```{r setup, message = FALSE}
library(shewhartr)
library(ggplot2)
library(dplyr)
```

Classical Shewhart charts assume the process mean is *stationary* —
constant in time, apart from chance variation. Many real processes
violate this assumption: a curing oven drifts, a microbial culture
grows, an epidemic curve rises and falls, sensor calibration shifts.
Applying a classical chart to a trended process produces systematically
wrong limits and a flood of false alarms (or, equivalently, false
reassurances).

The right approach is the one Mandel (1969) proposed in the *Journal
of Quality Technology*'s very first issue: fit a model for the trend
and place control limits around the *fitted curve*, not around a
constant centre line. This is what `shewhart_regression()` does.

## A simple example: linear drift

`temperature_drift` is 200 minutes of sensor readings on a curing
oven. The truth is a slow linear drift superimposed on a periodic
component plus noise.

```{r}
fit <- shewhart_regression(
  temperature_drift,
  value = temp_c,
  index = minute,
  model = "linear"
)
broom::glance(fit)
```

```{r}
autoplot(fit)
```

The chart's centre line follows the fitted line, and limits are
$\pm 3\hat{\sigma}_R$ where $\hat{\sigma}_R$ is estimated from the
moving range of the residuals.

## Sigma from residuals

A subtlety: the residuals from a regression fit are *correlated* under
ordinary least squares, even when the model is correctly specified
(adjacent residuals share the influence of the same fitted slope).
The classical $\overline{\mathrm{MR}}/d_2$ estimator partially absorbs
this, but for short series or autocorrelated noise we recommend
checking the residuals first via `shewhart_diagnostics()`:

```{r, eval = FALSE}
shewhart_diagnostics(fit)    # residuals~fitted, Q-Q, ACF, MR, histogram
```

If the ACF panel shows non-trivial autocorrelation, consider
modelling it explicitly (a wider topic; see Box, Jenkins & Reinsel
2008) before relying on the chart.

## Phase detection

`shewhart_regression()` can split the series into phases automatically:
when a runs rule fires (default Nelson 2 — nine consecutive points on
the same side of the fitted curve), a new phase begins and the model
is re-fit. This generalises the original v0.1.x behaviour with a
cleaner, configurable rule:

```{r}
set.seed(1)
trended <- tibble::tibble(
  t = 1:120,
  y = c( 1:60 * 0.5  + rnorm(60, sd = 0.5),     # phase 1
         30 + 1:60 * 0.1 + rnorm(60, sd = 0.5)) # phase 2: shift + slowdown
)

fit <- shewhart_regression(trended, value = y, index = t,
                           model = "linear",
                           phase_rule = "nelson_2_nine_same")
broom::glance(fit)
length(fit$fits)    # number of phases detected
```

The `phase_rule` argument accepts any rule from
`shewhart_rules_available()`. The legacy 7-points rule from `v0.1.x`
is still available as `"we_seven_same"`:

```{r}
fit_legacy <- shewhart_regression(trended, value = y, index = t,
                                  model = "linear",
                                  phase_rule = "we_seven_same")
length(fit_legacy$fits)
```

The trade-off is straightforward. With Nelson 2 (9 same side), the
in-control ARL is about 256 — false phase changes are rare. With the
WE 7-same rule, ARL_0 is about 64 — phase changes are detected
faster but at a higher false-alarm cost. See the
`arl-simulation` vignette for a quantitative comparison.

```{r}
autoplot(fit)            # Nelson 2 — usually 1–2 phases
autoplot(fit_legacy)     # WE 7 — typically more phases on the same data
```

## A worked example with many phases and visible violations

The COVID-19 mortality series for Recife is a textbook trended
process: a long, irregular climb, a peak, a slow decline. With a
single regression line and the legacy `we_seven_same` rule (the one
the original SBPO 2020 paper used), the chart partitions the series
into phases, fits a local trend in each, and flags the days when the
observed value departs sharply from the local trend.

```{r, fig.height = 4.6}
fit_recife <- shewhart_regression(
  cvd_recife,
  value      = new_deaths,
  index      = .t,
  model      = "loglog",
  phase_rule = "we_seven_same",
  rules      = c("nelson_1_beyond_3s", "we_seven_same"),
  lower_bound = 0          # death counts can't go negative
)

length(fit_recife$fits)              # phases detected
nrow(fit_recife$violations)          # individual flagged observations
autoplot(fit_recife)
```

Each shaded band is a phase. The dashed lines are that phase's
3-sigma limits, the solid line is the regression centre line for
that phase. Points highlighted in firebrick are the days flagged by
the rule set as departing from the local trend — these are the days
that motivated investigation in the original analysis.

The `violations` table tells you exactly which days fired which
rule:

```{r}
head(fit_recife$violations, 8)
```

## The model menu

| `model = ...`  | Functional form                                    |
|----------------|----------------------------------------------------|
| `"linear"`     | $y = \beta_0 + \beta_1 N$                          |
| `"log"`        | $\log(y + 1) = \beta_0 + \beta_1 N$                |
| `"loglog"`     | $\log(\log(y/\alpha + 1) + 1) = \beta_0 + \beta_1 N$ |
| `"gompertz"`   | Gompertz cumulative growth (via `nls`)             |
| `"logistic"`   | Logistic cumulative growth (via `nls`)             |
| `"auto"`       | Box-Cox guidance to choose between linear/log/log-log |

The `"auto"` setting calls `shewhart_box_cox()` internally and selects
based on the maximum-likelihood lambda. It is a good first try when
you don't have strong prior knowledge of the functional form.

For full control, supply your own formula:

```{r, eval = FALSE}
shewhart_regression(temperature_drift, value = temp_c, index = minute,
                    formula = temp_c ~ poly(minute, 3))
```

## A growth-curve example

`bacterial_growth` is a 24-hour OD time series whose true mean is a
Gompertz curve.

```{r}
fit_gomp <- shewhart_regression(
  bacterial_growth,
  value = od,
  index = hour,
  model = "gompertz"
)
broom::glance(fit_gomp)
```

```{r}
autoplot(fit_gomp)
```

The Gompertz parameterisation we use comes from Zwietering et al.
(1990) and is in `?Gompertz`.

## Interpreting violations

In a regression chart, a violation means an observation departs from
the *trend*, not from a constant baseline. This is exactly what we
want for trended processes: the question becomes "is the *deviation
from the expected trajectory* unusual?", not "is the value high or
low compared to a fixed reference?".

Phase changes are themselves interpreted as suspected shifts in the
underlying process — a re-tuned controller, a new operator, a new
batch of raw material. `shewhart_regression()` highlights them in
`autoplot()` by colouring each phase distinctly.

## References

* Mandel, B. J. (1969). The Regression Control Chart. *Journal of
  Quality Technology*, 1(1), 1-9.
* Hawkins, D. M. (1991). Multivariate Quality Control Based on
  Regression-Adjusted Variables. *Technometrics*, 33(1), 61-75.
* Wheeler, D. J., & Chambers, D. S. (1992). *Understanding Statistical
  Process Control* (2nd ed.). SPC Press.
* Box, G. E. P., Jenkins, G. M., & Reinsel, G. C. (2008). *Time Series
  Analysis: Forecasting and Control* (4th ed.). Wiley.
* Zwietering, M. H. et al. (1990). Modeling of the Bacterial Growth
  Curve. *Applied and Environmental Microbiology*, 56(6), 1875-1881.
* Box, G. E. P., & Cox, D. R. (1964). An Analysis of Transformations.
  *Journal of the Royal Statistical Society B*, 26(2), 211-252.
